The hydra and the rooster

### 2020-07-27 latex/chapter-ordinals.tex
redaction introduction to ordinals

Fortunately, the ordinals we need for  studying hydra battles are much simpler than Schütte's, and can be represented as rather simple data types in \gallina. So, we will use \emph{ordinal notations} (also called \emph{ordinal notation systems}). Let $\alpha$ be some countable ordinal; In \coq{} terms, an ordinal notation for $\alpha$ will be a data type for representing all ordinals below $\alpha$ (excluding $\alpha$) and a bunch of functions for comparing ordinals, doing some arithmetic, inspecting the topology of the order (limits, successors), and proofs of well-foundedness). These functions should be proved correct, with respect to Schütte's model.

@@

\subsection{Examples of Notation systems}

@@

\vspace{4pt}
\noindent\emph{From Library~\href{../src/html/hydras.OrdinalNotations/Omega_ON.html}{OrdinalNotations.Omega\_ON}}

@@

@@

In \coq{}, we use the disjoint union \texttt{(nat+nat)\%type} for representing the elements of $\omega+\omega$.
The blue ordinals of Figure~\ref{fig:omega-plus-omega} will be represented by terms of the form \texttt{(inl $i$)},
the red ones by \texttt{(inr $j$)}.

\vspace{4pt}
\noindent\emph{From Module~\href{../src/html/hydras.Prelude.Omega_plus_omega.html}{Prelude.Omega\_plus\_omega}}

\begin{Coqsrc}
Declare Scope oo_scope.
Delimit Scope oo_scope with oo.
Open Scope oo_scope.

Definition t := (nat + nat)%type.
Arguments inl  {A B} _.
Arguments inr  {A B} _.
\end{Coqsrc}

The order on type \texttt{t} is already defined in
Library~\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Relations.Relation_Operators.html}{Relations.Relation\_Operators}.

\begin{Coqsrc}
Definition lt : relation t := le_AsB _ _ Peano.lt Peano.lt.
Infix "<" := lt : oo_scope.

Definition le := clos_refl _ lt.

Infix "<=" := le : oo_scope.
\end{Coqsrc}

In order to build an instance of \texttt{ON}, we have to verify that \texttt{lt} is a well-founded strict order and define a correct comparison function.

\begin{Coqsrc}
Instance lt_strorder : StrictOrder lt.

Lemma lt_wf : well_founded lt.
Proof. apply wf_disjoint_sum; apply lt_wf. Qed.

Definition compare (alpha beta: t) : comparison :=
   match alpha, beta with
     inl _, inr _ => Lt
   | inl n, inl p | inr n, inr p => Nat.compare n p
   | inr _, inl _ => Gt
  end.

Lemma compare_correct alpha beta :
    CompareSpec (alpha = beta) (lt alpha beta) (lt beta alpha)
              (compare alpha beta).

Instance OmegaPlusOmega : ON lt compare.
Proof.
  split.
  - apply lt_strorder.
  - apply lt_wf.
  - apply compare_correct.
Qed.

@@

\subsection{Comparing Ordinal Notations}

>>>>

Fortunately, the ordinals we need for  studying hydra battles are much simpler than Schütte's, and can be represented as rather simple data types in \gallina. So, we will use \emph{ordinal notations} (also called \emph{[ordinal] notation systems}). Let $\alpha$ be some countable ordinal; In \coq{} terms, an ordinal notation for $\alpha$ will be a data type for representing all ordinals below $\alpha$ (excluding $\alpha$) and a bunch of functions for comparing ordinals, doing some arithmetic, inspecting the topology of the order (limits, successors), and proofs of well-foundedness). These functions should be proved correct, with respect to Schütte's model.

@@

\subsection{Instances of notation systems}

@@

\vspace{4pt}
\noindent\emph{From Library~\href{../src/html/hydras.OrdinalNotations/ON_Omega.html}{OrdinalNotations.ON\_Omega}}

@@

\subsection{Sum of two ordinal notations}

Let \texttt{NA} and \texttt{NB} be two ordinal notations, on the respective types \texttt{A} and \texttt{B}.

 We consider a new strict order
on the disjoint sum of the associated types, by putting all elements of \texttt{A} before the elements of \texttt{B} (thanks to Standard Library's relation operator \texttt{le\_AsB}).
(please look at Figure~\ref{fig:omega-plus-omega} for an example).

\emph{From Library~\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Relations.Relation_Operators.html}{Relations.Relation\_Operators}}.

\vspace{4pt}
\noindent\emph{From Library~\href{../src/html/hydras.OrdinalNotations/ON_plus.html}{OrdinalNotations.ON\_plus}}

\begin{Coqsrc}
Section Defs.

  Context `(ltA: relation A)
          (compareA : A -> A -> comparison)
          (NA: ON ltA compareA).
  Context `(ltB: relation B)
          (compareB : B -> B -> comparison)
          (NB: ON ltB compareB).

Definition t := (A + B)%type.
Arguments inl  {A B} _.
Arguments inr  {A B} _.

Definition lt : relation t := le_AsB _ _ ltA ltB.
\end{Coqsrc}

In order to build an instance of \texttt{ON}, we have to define a comparison function.

\begin{Coqsrc}
Definition compare (alpha beta: t) : comparison :=
   match alpha, beta with
     inl _, inr _ => Lt
   | inl a, inl a' => compareA a a'
   | inr b, inr b' => compareB b b'
   | inr _, inl _ => Gt
  end.

Lemma compare_correct alpha beta :
    CompareSpec (alpha = beta) (lt alpha beta) (lt beta alpha)
              (compare alpha beta).
\end{Coqsrc}

The Lemma \texttt{Wellfounded.Disjoint\_Union.wf\_disjoint\_sum} of Standard Library
helps us to prove that our order \texttt{lt} is well-founded.

\begin{Coqsrc}
Global Instance ON_plus : ON lt compare.
Proof.
  split.
  - apply lt_strorder.
  -  apply lt_wf.
  - apply compare_correct.
Qed.
\end{Coqsrc}

@@

We can define this notation in \coq{} as an instance of \texttt{ON\_plus}.

\vspace{4pt}
\noindent\emph{From Module~\href{../src/html/hydras.Prelude.ON_Omega_plus_omega.html}{Prelude.ON\_Omega\_plus\_omega}}

\begin{Coqsrc}
Definition Omega_plus_Omega := ON_plus Omega Omega.

Existing Instance Omega_plus_Omega.
Definition t := @ON_plus.t nat nat.
\end{Coqsrc}

\begin{Coqsrc}
Example ex1 : inl 7 < inr 0.
Proof. constructor. Qed.
\end{Coqsrc}

We can now define abbreviations. For instance, the finite ordinals will be represented by the terms built with  the constructor \texttt{inl}, and the first infinite ordinal $\omega$ by the term \texttt{(inr 0)}.

\begin{Coqsrc}
Definition fin (i:nat) : t := inl i.
Coercion fin : nat >-> t.

Notation "'omega'" := (inr  0:t).
\end{Coqsrc}

\begin{Coqsrc}
Example ex1' : fin 7 < omega.
Proof. constructor. Qed.
\end{Coqsrc}

\section{Limits and Successors}

Let us look again at our implementation of $\omega+\omega$. We can distinguish between the three kinds of ordinals seen in Fig~\ref{fig:ordinal-sequence}:

\begin{itemize}
\item The least ordinal, \texttt{(inl 0)}, also written \texttt{(fin 0)}.
\item The limit ordinal $\omega$.
\item The successor ordinals, either of the form \texttt{(inl (S $i$))} or \texttt{(inr (S $i$))}
\end{itemize}

It would be interesting to specify at the most generic level, what is a zero, a successor or a limit ordinal.

%%%% ICI ICI %%%%%%%%%%%

@@

\section{Comparing Ordinal Notations}
